Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Proofread through chapter 7

A method <{m}> with argument <{a}> is invoked in its
most basic form as follows (assigning the result to~\texttt{r}).
\begin{code}
<{r = m a}>
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
<{m(a)}>, the argument is (<{a}>), which is the same
as <{a}>.
If you invoke <{m()}>, the argument is <{()}>,
which is the empty tuple.
If you invoke <{m(a, b)}>, the argument is <{(a, b)}>,
the tuple consisting of values <{a}> and <{b}>.

@@

Harmony does not have a \textbf{return} statement.  (You can assign a return value
to a method by setting the <{result}> variable.)  Neither does it support
\textbf{break} or \textbf{continue} statements in loops.  One reason for their absence is
that, particularly in concurrent programming, such control flow directions are highly
error-prone.  It's too easy to forget to, say, release a lock when returning a value in the
middle of a method, a major source of bugs in practice.

Harmony is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If <{p}> is a pointer to a shared variable, then the
expression \texttt{!}<{p}> is the value of the shared variable.
In particular, <{!?x = x}>.
This is similar to how C pointers work (\texttt{*\string&}<{x}> = <{x}>).

@@

Finally, methods can have local variables.
Different from Python, Harmony local variables must be
explicitly bound within a scope.
Method variables are either mutable or immutable.
The arguments to a method and the bound variable (or
variables) within a <{for}> statement are immutable;
the variable <{result}> is mutable.
Using the <{var}> statement, new mutable local variables can be
declared.
For example, <{var}> $x = 3$ declares a new mutable local variable
<{x}>.
The <{let}> statement allows declaring new immutable local variables.
For example:
<{let}> $x = 3$: <{y}> $+$$=$ <{x}> adds 3 to the global variable
<{y}>.
See \autoref{sec:vars} for more information.

@@

If you are ready to learn about how locks are implemented in practice,
you can now skip the rest of this chapter.  But if you would like to
see a cool example of using the concepts introduced in this chapter,
hang on for a sequential Harmony program that finds anomalies in page
replacement algorithms.
In 1969, B\'{e}l\'{a}ady et al. published a paper~\cite{Belady69}
that showed that making a
cache larger does not necessarily lead to a higher hit ratio.  He showed
this for a scenario using a FIFO replacement policy when the cache is full. 
The program in \autoref{fig:clock} will find exactly the same scenario
if you define \texttt{FIFO} to be <{True}>.  Moreover, if you define
\texttt{FIFO} to be <{False}>, it will find a scenario for the
CLOCK replacement policy~\cite{clock}, often used in modern operating systems.

In this program, \texttt{CLOCK} maintains the state of a cache
(in practice, typically pages in memory).
The set \texttt{recent} maintains whether an access to the cache
for a particular reference was recent or not.
(It is not used if \texttt{FIFO} is <{True}>.)
The integer \texttt{misses} maintains the number of cache misses.
Method \texttt{ref}(\textit{ck}, <{x}>) is invoked when <{x}> is
referenced and checked against the cache \textit{ck}.

The program declares two caches: one with 3 entries (\textit{clock3})
and one with 4 entries (\textit{clock4}).
The interesting part is in the last block of code.
It runs every sequence of references of up to 100 entries, using
references in the range 1 through 5.
Note that all the constants chosen in this program (3, 4, 5, 100)
are the result of some experimentation---you can try other ones.
To reduce the search space, the first four references are pinned
to 1, 2, 3, and 4.
Further reducing the search space, the program never repeats the
same reference twice in a row (using the local variable \texttt{last}).

The two things to note here is the use of the \textbf{choose}
expression and the \textbf{assert} statement.  Using <{choose}>
we are able to express searching through all possible strings of
references without a complicated nested iteration.
Using \texttt{assert} we are able to express the anomaly we are
looking for.

In case you want to check if you get the right results.  For \texttt{FIFO},
the program finds (in a few minutes) the same anomaly that
B\'{e}l\'{a}dy et al. found: 1 2 3 4 1 2 5 1 2 3 4 5.
For the \texttt{CLOCK} algorithm the program actually finds a shorter
reference string: 1 2 3 4 2 1 2 5 1 2.

>>>>

A method <{m}> with argument <{a}> is invoked in its
most basic form as follows (assigning the result to~<{r}>).
\begin{code}
<{r = m a}>
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
<{m(a)}>, the argument is (<{a}>), which is the same
as <{a}>.
If you invoke <{m()}>, the argument is <{()}>,
which is the empty tuple.
If you invoke <{m(a, b)}>, the argument is <{(a, b)}>,
the tuple consisting of values <{a}> and <{b}>.

@@

Harmony does not have a \textbf{return} statement.  (You can assign a return value
to a method by setting the <{result}> variable.)  Neither does it support
\textbf{break} or \textbf{continue} statements in loops.  One reason for their absence is
that, particularly in concurrent programming, such control flow directions are highly
error-prone.  It's too easy to forget to, say, release a lock when returning a value in the
middle of a method---a major source of bugs in practice.

Harmony is not an object-oriented language like Python is.  In Python,
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If <{p}> is a pointer to a shared variable, then the
expression \texttt{!}<{p}> is the value of the shared variable.
In particular, <{!?x = x}>.
This is similar to how C pointers work (\texttt{*\string&}<{x}> = <{x}>).

@@

Finally, methods can have local variables.
Different from Python, Harmony local variables must be
explicitly bound within a scope.
Method variables are either mutable (writable) or immutable (read-only).
The arguments to a method and the bound variable (or
variables) within a <{for}> statement are immutable;
the variable <{result}> is mutable.
Using the <{var}> statement, new mutable local variables can be
declared.
For example, <{var}> $x = 3$ declares a new mutable local variable
<{x}>.
The <{let}> statement allows declaring new immutable local variables.
For example:
<{let}> $x = 3$: <{y}> $+$$=$ <{x}> adds 3 to the global variable
<{y}>.
See \autoref{sec:vars} for more information.

@@

If you are ready to learn about how locks are implemented in practice,
you can now skip the rest of this chapter.  But if you would like to
see a cool example of using the concepts introduced in this chapter,
hang on for a sequential Harmony program that finds anomalies in page
replacement algorithms.
In 1969, B\'{e}l\'{a}dy et al.~published a paper~\cite{Belady69}
that showed that making a
cache larger does not necessarily lead to a higher hit ratio.  He showed
this for a scenario using a FIFO replacement policy when the cache is full. 
The program in \autoref{fig:clock} will find exactly the same scenario
if you define <{FIFO}> to be <{True}>.  Moreover, if you define
<{FIFO}> to be <{False}>, it will find a scenario for the
CLOCK replacement policy~\cite{clock}, often used in modern operating systems.

In this program, <{CLOCK}> maintains the state of a cache
(in practice, typically pages in memory).
The set <{recent}> maintains whether an access to the cache
for a particular reference was recent or not.
(It is not used if <{FIFO}> is <{True}>.)
The integer <{misses}> maintains the number of cache misses.
Method <{ref(ck, x)}> is invoked when <{x}> is
referenced and checked against the cache <{ck}>.

The program declares two caches: one with 3 entries (<{clock3}>)
and one with 4 entries (<{clock4}>).
The interesting part is in the last block of code.
It runs every sequence of references of up to 100 entries, using
references in the range 1 through 5.
Note that all the constants chosen in this program (3, 4, 5, 100)
are the result of some experimentation---you can try other ones.
To reduce the search space, the first four references are pinned
to 1, 2, 3, and 4.
Further reducing the search space, the program never repeats the
same reference twice in a row (using the local variable <{last}>).

The two things to note here is the use of the \textbf{choose}
expression and the \textbf{assert} statement.  Using <{choose}>,
we are able to express searching through all possible strings of
references without a complicated nested iteration.
Using \texttt{assert}, we are able to express the anomaly we are
looking for.

In case you want to check if you get the right results.  For <{FIFO}>,
the program finds the same anomaly that
B\'{e}l\'{a}dy et al.~found: 1 2 3 4 1 2 5 1 2 3 4 5.
For the <{CLOCK}> algorithm the program actually finds a shorter
reference string: 1 2 3 4 2 1 2 5 1 2.
